Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models

### 2015-02-21 processes-69103_review_response/processes-69103.tex
Some extra changes

Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Khanin:1989aa,Willems:1991aa,Baldwin:1994aa,Leipold:1995aa,Kuharsky:2001aa}.
Mechanistic ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, Hockin \emph{et al.} \citep{Hockin:2002aa}, and later Butenas et al., \citep{Butenas:2004aa}
who developed and then subsequently refined highly mechanistic coagulation models. Recently Papadopoulos and co-workers used a phenomenological  mathematical model for thrombin generation \cite{Atkin:2014}. Using a set of four ordinary differential equations they were able to derive an equation for temporal evolution of thrombin generation. However the model focuses more on thrombin generation rather than dynamics of coagulation. Unlike our hybrid approach it does not consider regulatory detail and thus is unable to capture the roles of primary inhibitors like ATIII or protein C. Thus in cases where we need to understand the dynamics of coagulation without complex modeling or when there is incomplete mechanistic knowledge, the hybrid approach is a better candidate. The model also requires that the parameters are adjusted or fine tuned for each data set unlike the hybrid approach where we use a single ensemble across all experiments. However platelet modeling is a significant advantage of this model over the hybrid approach.  Future versions of the hybrid model can involve modeling platelets as done by Papdopoulos et al. in their phenomenological model. Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa}, and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
However, these previous studies were largely based upon extensive mechanistic knowledge. 
This is possible because blood, while enormously complex, can be systematically interrogated.
Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which a complete mechanistic understanding is missing. 
We tested this approach by modeling the human coagulation cascade. 
The hybrid model consisted of only five differential equations augmented with several logical rules.
Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

Malfunctions in coagulation can have potentially fatal consequences.
Aggressive clotting involved with Coronary Artery Diseases (CADs), collectively accounts for 38\% of all deaths in North America \cite{HANSSON2005}.
Coagulation management during surgery can also be challenging, particularly with the increase in clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
The coagulation factors VIII (fVIII) and IX (fIX) are deficient in Hemophilia A and B, respectively \cite{COOPER1994,MANNUCCI2001,MITCHELL2008}.
People with mild hemophilia have 5-40\% of the normal clotting factor levels while severe hemophiliacs have $<$1\% \cite{MITCHELL2008}.
Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
However, clotting factor replacement sometimes leads to the formation of fVIII and fIX inhibitors \emph{in vivo} \cite{TOMOKIYO2003}.
Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
hemophilia with and without factor VIII/IX inhibitors \cite{Shapiro:2008px}.
However, rFVIIa requires frequent administration (every 2-3 hr), and many questions remain about its mechanism of action, 
its effective dosage \cite{TOMOKIYO2003}, and its overall utility for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. 
In this study, we did not model rFVIIa-induced coagulation directly. 
Rather, we modeled a general trigger which initiated the extrinsic coagulation cascade. 
Since we identified the model using TF/FVIIa, inherent to our rFVIIa simulations (and the rate constant governing initiation) 
was the presence of TF. However, even with this complication, 
the model generated potentially useful insight into the rFVIIa mechanism of action, and its possible shortcomings especially for the treatment of hemophilia. 
The addition of rFVIIa directly activated thrombin through the initiation pathway. 
However, no amplification of the thrombin signal occurred without fVIII or fIX. 
Thus, the peak thrombin signal was lower than normal coagulation, the peak thrombin time was longer, and thrombin generation was eventually inhibited by the combined action of ATIII and the protein C pathway.
However, as the dose of rFVIIa increased, the peak thrombin time decreased (eventually saturating around 200$\times$nominal trigger), and the peak thrombin value increased such that
the thrombin profile resembled normal coagulation.
Butenas \emph{et al.} performed an extensive \textit{in vitro} study of rFVIIa-induced thrombin generation under normal and hemophilic conditions \citep{Butenas:2002ab}.
They found qualitatively similar trends, namely rFVIIa restored normal coagulation (even in the absence of TF) for large enough rFVIIa doses, although rFVIIa-induced coagulation 
in hemophilia (even for large rFVIIa doses) lagged the normal profile. 
These results suggest that rFVIIa administration alone
might not be able to initiate normal coagulation in recurrent bleeding, unless the dosage is well above a critical threshold.
However, defining this threshold, which is likely patient specific, is difficult as there is tremendous patient to patient variability even with a normal
coagulation phenotype \citep{Danforth:2012aa}.

A number of mathematical models and computational simulations concerning coagulation dynamics have been based on the Hockin-Mann model of coagulation.

@@

>>>>

Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Khanin:1989aa,Willems:1991aa,Baldwin:1994aa,Leipold:1995aa,Kuharsky:2001aa}.
Mechanistic ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, Hockin \emph{et al.} \citep{Hockin:2002aa}, and later Butenas et al., \citep{Butenas:2004aa}
who developed and then subsequently refined highly mechanistic coagulation models. Recently Papadopoulos and co-workers used a phenomenological  mathematical model for thrombin generation \cite{Atkin:2014}. Using a set of four ordinary differential equations they were able to derive an equation for temporal evolution of thrombin generation. However the model has a greater focus on thrombin generation and lesser emphasis on the influence of various regulatory elements on the dynamics of coagulation. Unlike our hybrid approach it does not model regulatory detail and thus may be unable to capture the roles of primary inhibitors like ATIII or protein C. Thus in cases where we need to understand the dynamics of coagulation without complex modeling or when there is incomplete mechanistic knowledge, the hybrid approach is a better candidate. The model also requires that the parameters are adjusted or fine tuned for each data set unlike the hybrid approach where we use a single ensemble across all experiments. However platelet modeling is a significant advantage of this model over the hybrid approach.  Future versions of the hybrid model can involve modeling platelets as done by Papdopoulos \emph{et al.}, in their phenomenological model. Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa}, and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
However, these previous studies were largely based upon extensive mechanistic knowledge. 
This is possible because blood, while enormously complex, can be systematically interrogated.
Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which a complete mechanistic understanding is missing. 
We tested this approach by modeling the human coagulation cascade. 
The hybrid model consisted of only five differential equations augmented with several logical rules.
Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

Malfunctions in coagulation can have potentially fatal consequences.
Aggressive clotting involved with Coronary Artery Diseases (CADs), collectively accounts for 38\% of all deaths in North America \cite{HANSSON2005}.
Coagulation management during surgery can also be challenging, particularly with the increase in clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
The coagulation factors VIII (fVIII) and IX (fIX) are deficient in Hemophilia A and B, respectively \cite{COOPER1994,MANNUCCI2001,MITCHELL2008}.
People with mild hemophilia have 5-40\% of the normal clotting factor levels while severe hemophiliacs have $<$1\% \cite{MITCHELL2008}.
Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
However, clotting factor replacement sometimes leads to the formation of fVIII and fIX inhibitors \emph{in vivo} \cite{TOMOKIYO2003}.
Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
hemophilia with and without factor VIII/IX inhibitors \cite{Shapiro:2008px}.
However, rFVIIa requires frequent administration (every 2-3 hr), and many questions remain about its mechanism of action, 
its effective dosage \cite{TOMOKIYO2003}, and its overall utility for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. 
In this study, we did not model rFVIIa-induced coagulation directly. 
Rather, we modeled a general trigger which initiated the extrinsic coagulation cascade. 
Since we identified the model using TF/FVIIa, inherent to our rFVIIa simulations (and the rate constant governing initiation) 
was the presence of TF. However, even with this complication, 
the model generated potentially useful insight into the rFVIIa mechanism of action, and its possible shortcomings especially for the treatment of hemophilia. 
The addition of rFVIIa directly activated thrombin through the initiation pathway. 
However, no amplification of the thrombin signal occurred without fVIII or fIX. 
Thus, the peak thrombin signal was lower than normal coagulation, the peak thrombin time was longer, and thrombin generation was eventually inhibited by the combined action of ATIII and the protein C pathway.
However, as the dose of rFVIIa increased, the peak thrombin time decreased (eventually saturating around 200$\times$nominal trigger), and the peak thrombin value increased such that
the thrombin profile resembled normal coagulation.
Butenas \emph{et al.} performed an extensive \textit{in vitro} study of rFVIIa-induced thrombin generation under normal and hemophilic conditions \citep{Butenas:2002ab}.
They found qualitatively similar trends, namely rFVIIa restored normal coagulation (even in the absence of TF) for large enough rFVIIa doses, although rFVIIa-induced coagulation 
in hemophilia (even for large rFVIIa doses) lagged the normal profile. 
These results suggest that rFVIIa administration alone
might not be able to initiate normal coagulation in recurrent bleeding, unless the dosage is well above a critical threshold.
However, defining this threshold, which is likely patient specific, is difficult as there is tremendous patient to patient variability even with a normal
co agulation phenotype \citep{Danforth:2012aa}.

A number of mathematical models and computational simulations concerning coagulation dynamics have been based on the Hockin-Mann model of coagulation. Recently, Brummel-Ziedins and co-workers built a model that incorporated the Protein C (PC) pathway into the Hockin-Mann network to investigate thrombin generation in cases of familial PC deficiency \cite{BrummelPC:2012}. Computational simulations using this model showed that PC mutation caused an elevation in peak thrombin levels without changing the initiation time or the slope of thrombin generation curve. Our hybrid model to a certain extent captured this trend qualitatively. We were able to show the decrease in peak thrombin concentration in the presence of PC pathway without any other inhibitors like ATIII. Danforth \emph{et al.} computationally simulated normal thrombin generation using 5pM tissue factor with all other factors at their mean physiological level \citep{Danforth:2012aa}. The initiation time in this simulation was approximately 4.4 minutes. When predicting the normal thrombin generation curve using 0.2 nM of FVIIa we were able to show an approximate initiation time around 5 minutes with a similar thrombin generation profile (Fig. \ref{fig-hemophilia}A).  In a kinetic modeling to study the effect of rfVIIa on thrombin generation \cite{Mitrophanov:2011} it was shown that increasing amounts of FVIIa accelerated the initiation phase and the propagation phase during thrombin generation. Analysis of thrombin generation curves showed that the FVIIa controlled the maximal slope of thrombin curve along with peak thrombin time and initiation time. Though we had issues in capturing the effect of FVIIa on initiation time we were able to show that change in concentration of FVIIa affected the maximal thrombin slope as well as propagation phase (Fig. \ref{fig-rFVII}A).

@@

\bibitem{BrummelPC:2012}
Kathleen E. Brummel-Ziedins,Thomas Orfeo,Peter W. Callas, Matthew Gissel, Kenneth G. Mann, Edwin G. Bovill (2012) The Prothrombotic Phenotypes in Familial Protein C Deficiency Are Differentiated by Computational Modeling of Thrombin Generation
\newblock PLoS One 5: e44378.

\bibitem{Mitrophanov:2011}
Alexander Y. Mitrophanov, Jaques Reifman(2011) Kinetic modeling sheds light on the mode of action of recombinant factor VIIa on thrombin generation
\newblock Thrombosis Research Volume 128, Issue 4, October 2011, Pages 381-390.
